---
title: "Reverse Engineering Project"
author: "Student names here"
date: "Date here"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [STORY NAME](STORY URL)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
options(scipen=999)
library(tidyverse)
library(lubridate)
library(janitor)
library(dplyr)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 

SportsDonations_sports_political_donations <- read_csv("SportsDonations - sports-political-donations.csv")
```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: "Overall, 51.7 percent of all contributions by WNBA owners went to Democrats, with 42.3 percent going to Republicans."
* **Analysis summary**: What I actually found when I ran this code was nearly the opposite of this sentence. I found that about 51.7% of the contributions from WNBA owners went to Republicans, with about 43.8% going to Democrats. Then there were other minor percentages that went to bipartisan groups.

```{r}
# Put code to reverse engineer sentence here
SportsDonations_sports_political_donations <- SportsDonations_sports_political_donations |>
  clean_names()
wnba_owners <- SportsDonations_sports_political_donations |>
  filter(league == "WNBA") |>
  group_by(party) |>
  summarize(Total_Amount = sum(amount, na.rm = TRUE)) |>
  mutate(Percentage = (Total_Amount / sum(Total_Amount)) * 100)

# Display results of code below this codeblock
wnba_owners
```

### Sentence 2

* **Sentence text**: "Thirteen owners have contributed directly to Trump’s campaign or Trump super PACs in the 2020 election cycle, with contributions adding up to $1,746,423."
* **Analysis summary**: My analysis yielded a very similar result, but it wasn't exactly matching. My analysis showed that it was 12 unique owners, and the contributions added up to $1,544,421.

```{r}
# Put code to reverse engineer sentence here
SportsDonations_sports_political_donations <- SportsDonations_sports_political_donations |>
  clean_names()
trump_keywords <- c("Trump", "Donald J. Trump", "Donald J Trump", "Make America Great Again", 
                    "Donald Trump for President", "Trump Victory", "Save America", "Trump MAGA Committee")
trump_donations_2020 <- SportsDonations_sports_political_donations |>
  filter(election_year == 2020) |>
  filter(str_detect(tolower(recipient), 
                    paste(tolower(trump_keywords), collapse = "|"))) |>
  filter(!is.na(owner)) 
trump_contributions_summary <- trump_donations_2020 |>
  summarize(
    unique_owners = n_distinct(owner),
    total_contributed = sum(amount, na.rm = TRUE)
  )
# Display results of code below this codeblock
trump_contributions_summary
```

### Sentence 3

* **Sentence text**: "By contrast, 20 owners have contributed directly to Biden or Biden’s super PACs, though at smaller totals, with overall contributions equaling $405,745."
* **Analysis summary**: This analysis also yielded pretty similar results, but my code found that 21 owners contributed directly, with overall contributions adding up to $308,545.

```{r}
# Put code to reverse engineer sentence here
biden_keywords <- c("Biden", "Joe Biden", "Joseph R. Biden", "Biden for President", 
                    "Unite the Country", "Biden Victory Fund", "Future Forward USA")
biden_donations_2020 <- SportsDonations_sports_political_donations |>
  filter(election_year == 2020) |>
  filter(str_detect(tolower(recipient), 
                    paste(tolower(biden_keywords), collapse = "|"))) |>
  filter(!is.na(owner))
biden_contributions_summary <- biden_donations_2020 |>
  summarize(
    unique_owners = n_distinct(owner),
    total_contributed = sum(amount, na.rm = TRUE)
  )
# Display results of code below this codeblock
biden_contributions_summary
```

### Sentence 4

* **Sentence text**: "A relatively small group of donors makes up a huge chunk of the overall donation data. Forty owners account for more than $40 million — more than 85 percent of all contributions."
* **Analysis summary**: This analysis below basically matches exactly what the text above says. The top 40 contributing owners spend more than 41.5 million dollars, and account for 89% of the total spending.

```{r}
# Put code to reverse engineer sentence here
SportsDonations_sports_political_donations <- SportsDonations_sports_political_donations |>
  clean_names()
owner_totals <- SportsDonations_sports_political_donations |>
  filter(!is.na(owner)) |>
  group_by(owner) |>
  summarize(owner_total = sum(amount, na.rm = TRUE)) |>
  arrange(desc(owner_total))
top_40_owners <- owner_totals |>
  slice_max(order_by = owner_total, n = 40)
top_40_total <- sum(top_40_owners$owner_total, na.rm = TRUE)
all_owners_total <- sum(owner_totals$owner_total, na.rm = TRUE)
percent_top_40 <- (top_40_total / all_owners_total) * 100
# Display results of code below this codeblock
tibble(
  top_40_total = top_40_total,
  all_owners_total = all_owners_total,
  percent_top_40 = percent_top_40
)
```

### Sentence 5

* **Sentence text**: "And looking specifically at these big spenders, it’s clear where their allegiances lie: 77.7 percent of that $40 million went to Republican causes or candidates, with 18.8 percent going to Democrats."
* **Analysis summary**: This analysis that I ran below is pretty close to matching the sentence above. I found that about 20.79% of the 40 million dollars went to Democrats and about 75.8% went to Republicans. This data could be a bit off depending on what the article counted as belonging to the various bipartisan groups.

```{r}
# Put code to reverse engineer sentence here
top_40_owner_names <- top_40_owners$owner
top_40_donations <- SportsDonations_sports_political_donations |>
  filter(owner %in% top_40_owner_names)
top_40_party_breakdown <- top_40_donations |>
  group_by(party) |>
  summarize(total = sum(amount, na.rm = TRUE)) |>
  mutate(percentage = (total / sum(total)) * 100)
# Display results of code below this codeblock
top_40_party_breakdown
```

-30-