---
title: "Reverse Engineering Project"
author: "Student names here"
date: "Date here"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [STORY NAME](STORY URL)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
options(scipen=999)
library(tidyverse)
library(lubridate)
library(janitor)
library(dplyr)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 

SportsDonations_sports_political_donations <- read_csv("SportsDonations - sports-political-donations.csv")
```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: "Overall, 51.7 percent of all contributions by WNBA owners went to Democrats, with 42.3 percent going to Republicans."
* **Analysis summary**: What I actually found when I ran this code was nearly the opposite of this sentence. I found that about 51.7% of the contributions from WNBA owners went to Republicans, with about 43.8% going to Democrats. Then there were other minor percentages that went to bipartisans. 

```{r}
# Put code to reverse engineer sentence here
SportsDonations_sports_political_donations <- SportsDonations_sports_political_donations |>
  clean_names()
wnba_owners <- SportsDonations_sports_political_donations |>
  filter(league == "WNBA") |>
  group_by(party) |>
  summarise(Total_Amount = sum(amount, na.rm = TRUE)) |>
  mutate(Percentage = (Total_Amount / sum(Total_Amount)) * 100)

# Display results of code below this codeblock

```

### Sentence 2

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 3

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 4

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 5

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-