---
title: "FinalDataAssignment"
author: "Ryan Alonardo"
date: "2025-05-08"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
df <- play_off_box_scores_2010_2024
```

For this question, I just wanted to figure out which players were the most dominant in a playoff run; basically yielding the most memorable runs by a player in the last 14 years. This data matched up with what I figured it would look like, showing that the data is in fact accurate and works, and also showed who's had a huge impact on the sport.

```{r}
#Q1. Which players have had the greatest impact in a playoff run?

player_impact_summary <- df |>
  group_by(season_year, teamSlug, personName) |>
  summarize(
    total_points = sum(points, na.rm = TRUE),
    total_assists = sum(assists, na.rm = TRUE),
    total_rebounds = sum(reboundsTotal, na.rm = TRUE),
    total_plus_minus = sum(plusMinusPoints, na.rm = TRUE),
    games_played = n(),
    .groups = "drop"
  ) |>

  mutate(
    impact_score = total_points + total_assists * 1.5 + total_rebounds * 1.2 + total_plus_minus * 1.5
  )

ranked_players <- player_impact_summary |>
  group_by(season_year, teamSlug) |>
  arrange(desc(impact_score)) |>
  mutate(team_rank = row_number()) |>
  ungroup()

top_3_per_team <- ranked_players |>
  filter(team_rank <= 3)

print(top_3_per_team)

top_player_2024 <- ranked_players |>
  filter(season_year == 2024, team_rank == 1)
```

In this question, I wanted to go about figuring out which players had the biggest drop off in terms of production in consecutive playoff years. I noticed for this question that some players that were listed actually sustained injuries the following postseason, which led to the lack of production, but some simply did not make it as far in the postseason, explaining their lack of points, rebounds, assists, etc.

```{r}

#Q2. Which players had the biggest playoff drop-offs (consecutive years)?

library(dplyr)
library(ggplot2)

df <- play_off_box_scores_2010_2024 |>`
  mutate(minutes = as.numeric(minutes))  

seasonal_performance <- df |>
  group_by(personName, season_year) |>
  summarize(
    total_minutes = sum(minutes, na.rm = TRUE),
    total_points = sum(points, na.rm = TRUE),
    total_rebounds = sum(reboundsTotal, na.rm = TRUE),
    total_assists = sum(assists, na.rm = TRUE),
    total_steals = sum(steals, na.rm = TRUE),
    total_blocks = sum(blocks, na.rm = TRUE),
    total_turnovers = sum(turnovers, na.rm = TRUE),
    games_played = n(),
    .groups = "drop"
  ) |>
  filter(total_minutes >= 50) |>  
  mutate(
    performance_score = total_points + total_rebounds + total_assists +
                        total_steals + total_blocks - total_turnovers
  )

drop_offs <- seasonal_performance |>
  arrange(personName, season_year) |>
  group_by(personName) |>
  mutate(
    next_year = lead(season_year),
    next_score = lead(performance_score)
  ) |>
  filter(!is.na(next_score)) |> 
  mutate(
    dropoff = performance_score - next_score
  ) |>
  ungroup()

biggest_dropoffs <- drop_offs |>
  arrange(desc(dropoff)) |>
  head(15)

print(biggest_dropoffs)
```

In this question, I looked at which team has the most ball-dominant, heliocentric players and offenses in the postseason. From this, I was able to compare the top 10 players and teams to whether or not they had success, and was able to see if a ball-dominant player or a more team-oriented offense yielded better postseason results.

```{r}
#Q3. Which teams rely most heavily on a single player during the playoffs?

library(dplyr)
library(ggplot2)

df <- play_off_box_scores_2010_2024 |>
  mutate(minutes = as.numeric(minutes))

player_team_stats <- df |>
  group_by(season_year, teamSlug, personName) |>
  summarize(
    total_points = sum(points, na.rm = TRUE),
    total_assists = sum(assists, na.rm = TRUE),
    total_minutes = sum(minutes, na.rm = TRUE),
    .groups = "drop"
  )

team_totals <- player_team_stats |>
  group_by(season_year, teamSlug) |>
  summarize(
    team_points = sum(total_points, na.rm = TRUE),
    team_assists = sum(total_assists, na.rm = TRUE),
    team_minutes = sum(total_minutes, na.rm = TRUE),
    .groups = "drop"
  )

player_share <- player_team_stats |>
  inner_join(team_totals, by = c("season_year", "teamSlug")) |>
  mutate(
    point_share = total_points / team_points,
    assist_share = total_assists / team_assists,
    minutes_share = total_minutes / team_minutes,
    reliance_score = point_share * 0.5 + assist_share * 0.3 + minutes_share * 0.2
  )

most_reliant_players <- player_share |>
  group_by(season_year, teamSlug) |>
  arrange(desc(reliance_score)) |>
  slice(1) |>
  ungroup()

top_reliant_teams <- most_reliant_players |>
  arrange(desc(reliance_score)) |>
  select(season_year, teamSlug, personName, point_share, assist_share, minutes_share, reliance_score)

print(head(top_reliant_teams, 10))
```

In this question, I looked at some of the all-time single game performances by players from an equation model I made. This was able to tell me which players were able to step up in specific pivotal momemnts in various playoff runs, shaping who we might consider as clutch players.

```{r}
#4. What are the most well-rounded (all-around) playoff single-game performances from 2010–2024?

library(dplyr)
library(ggplot2)

df <- play_off_box_scores_2010_2024 |>
  mutate(minutes = as.numeric(minutes))  

well_rounded_games <- df |>
  mutate(
    well_rounded_score = points +
      reboundsTotal +
      assists +
      1.5 * steals +
      1.5 * blocks -
      turnovers
  ) |>
  select(season_year, game_date, teamSlug, personName, minutes, points, reboundsTotal,
         assists, steals, blocks, turnovers, well_rounded_score)

well_rounded_filtered <- well_rounded_games |>
  filter(minutes >= 20) 

top_well_rounded <- well_rounded_filtered |>
  arrange(desc(well_rounded_score)) |>
  head(20)

print(top_well_rounded)
```

Finally, this question looked at the old adage of "defense wins championships," and I wanted to test that. From this question, I was able to compare the top 10 names to whether or not that team had postseason success or won a championship, to see if the saying held true, or if times were changing.

```{r}
#5. Which players are defensive anchors in the playoffs?

library(dplyr)
library(ggplot2)

df <- play_off_box_scores_2010_2024 |>
  mutate(minutes = as.numeric(minutes)) 

defensive_summary <- df |>
  group_by(personName, season_year) |>
  summarize(
    total_minutes = sum(minutes, na.rm = TRUE),
    total_def_rebounds = sum(reboundsDefensive, na.rm = TRUE),
    total_steals = sum(steals, na.rm = TRUE),
    total_blocks = sum(blocks, na.rm = TRUE),
    total_plus_minus = sum(plusMinusPoints, na.rm = TRUE),
    games_played = n(),
    .groups = "drop"
  ) |>
  filter(total_minutes >= 100) |>
  
  mutate(
    defensive_impact_score = total_blocks * 2 +
      total_steals * 1.5 +
      total_def_rebounds * 1.2 +
      total_plus_minus
  )

top_defenders <- defensive_summary |>
  arrange(desc(defensive_impact_score)) |>
  head(20)

print(top_defenders)
```