---
title: "lab_11"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Our usual libraries for working with data, plus rvest and lubridate.

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
library(tidyverse)  
library(rvest)     
library(lubridate)   
library(janitor)    
library(glue)       

options(stringsAsFactors = FALSE)  
```

Let's get to scraping.

## Questions

**Q1**. Scrape the listing of available Maryland physician disciplinary alerts at <https://www.mbp.state.md.us/disciplinary_2024.aspx> into a dataframe. You should have three columns, one of which is a date, so make sure the date column has a date datatype. What's the most common sanction, and how many times is it mentioned in the 2024 alerts?

**A1** The most common sanction was "Summary Suspension Affirmed" and "Surrender of License" and they were mentioned seven times.

```{r}
url <- "https://www.mbp.state.md.us/disciplinary_2024.aspx"

webpage <- read_html(url)

disciplinary_data <- html_node(webpage, "table") |>
  html_table(fill = TRUE) |>
  clean_names() |>
  mutate(date = mdy(date))  

most_common_sanction <- disciplinary_data |>
  group_by(sanction) |>
  summarize(count = n()) |>
  arrange(desc(count))

most_common_sanction
```

**Q2** Next, let's scrape the list of press releases from Maryland's Office of the Public Defender, <https://www.opd.state.md.us/press-releases>. This isn't a table, so you'll need to use `html_elements()` and your browser's inspector and do some clean up on the results. The result should be a dataframe with two columns that contain the date and title, and the date column should have a date datatype. The challenge here is figuring out how to isolate the releases.

When you finish scraping into a dataframe, write code to find the press releases that have the word "police" in the title.

**A2** The number of press releases that have "police" in the title is 66.

```{r}
md_press_release_url <- "https://www.opd.state.md.us/press-releases"

md_press_release_data <- md_press_release_url |>
  read_html() |>
  html_elements('p.wixui-rich-text__text span a') |>
  html_text() |>
  as_tibble()
md_press_release_data

print(head(press_releases))

police_releases <- press_releases |>
  filter(str_detect(str_to_lower(title), "police"))

print(police_releases)
```

**Q3** Sen. Chris Van Hollen, D-Maryland, has posted press releases at <https://www.vanhollen.senate.gov/news/press-releases>. It would be great to have them in a dataframe that has the following columns: date, title and url.

To do this, you will need to scrape the page's html and save that to a variable, and *then* extract the dates, titles and urls into separate dataframes using html_elements(). There are two different elements that contain the press release URL; pick one of them, and you may need to invoke an HTML `class` value to do it. The function `html_text()` pulls out the contents of a tag, but for urls we want the HTML attribute. Rvest gives you a way to extract the URL from a link; google to find out what it is. Remember how we turn a list into a dataframe.

At the end, you'll have three dataframes that you want to combine into a single dataframe. When we want to combine the rows of identical dataframes, we used `bind_rows()`. If you were combining columns instead of rows, there's a similar function. Find out what it is and use it to put all of the dataframes together into a single one.

When you're done, rename the columns so they make sense, then make sure the date column is an actual date.

Finally, tell me what questions you could ask of this data if you had all of his press releases. Be creative.

**A3** Some of the possible questions I could explore of his press conferences include: What is the sentiment of his press conferences? Which keywords or phrases are most popular? How do the topics of his press conferences change over time?

```{r}
vhollen_releases_url <- "https://www.vanhollen.senate.gov/news/press-releases"

vhollen_html <- vhollen_releases_url |>
  read_html()

vhollen_releases_date <- vhollen_html |>
  html_elements('.ArticleBlock__titleContainer .Heading--time') |>
  html_text() |>
  as_tibble() |>
  rename(date = value)
vhollen_releases_date

vhollen_releases_title <- vhollen_html |>
  html_elements('.ArticleBlock__titleContainer h2') |>
  html_text() |>
  as_tibble() |>
  rename(title = value)
vhollen_releases_title

vhollen_releases_url <- vhollen_html |>
  html_elements('.ArticleBlock__titleContainer a') |>
  html_attr("href") |>
  as_tibble() |>
  rename(url = value)
vhollen_releases_url

vhollen_data <-
  bind_cols(vhollen_releases_date, vhollen_releases_title, vhollen_releases_url)
vhollen_data

cleaned_vhollen_data <-
  vhollen_data |>
  mutate(date=mdy(date))
cleaned_vhollen_data
```
