---
title: "lab_01"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you? **A1**
The Computer, Math and Natural Sciences department has the highest mean number of seats for its classes. What's most interesting is that it has just one class in the department. Departments like Engineering Science with 809 classes and a mean of 127 seats seem more significant.
```{r} 
umd_courses |>
  group_by(department) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats),
    mean_seats = mean(seats),
    median_seats = median(seats),
    minimum_seats = (min(seats)),
    maximum_seats = (max(seats)),
  ) |>
  arrange(desc(mean_seats)) 
```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. Your code must show the greatest number of classes first.
**A2**
AASP100 has been offered 29 times.
```{r}
umd_courses |>
  count(id, sort = TRUE) |>
  slice_head(n = 1)
```

**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most?
**A3**
The instructor has not come up when I run this particular set of code, only NA, which seems to be a decent amount of the instructors in this dataset.
```{r}
umd_courses |>
  count(instructors, department, sort = TRUE) |>
  slice_head(n = 1)
```

**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why? **A4**

I think I would have gotten a more accurate answer if I was able to filter out every single "NA" response for the instructor variable so I could only focus on the actual names of instructors. I'm not exactly sure how to run this code, but I know it would only leave me with the names of instructors. From there, I'm able to find the mode instructor name, or the one that shows up the most.

**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? **A5**

Below, I inserted the code I used to figure out my second question. I wanted to figure out what some of the interesting courses were in the School of Music. Some of them are quite interesting, such as "Opera Techniques" and "Aural Skills for Conductors."
```{r}
school_of_music_courses <- umd_courses |>
  filter(department == "School of Music")
print(school_of_music_courses)
```
